### Code was adapted from https://github.com/Capblancq/RDA-landscape-genomics/blob/e2cb3cb5530e286d79875f619314d563938145e3/RDA_landscape_genomics.Rmd
---
title: "Adaptive Indices"
author: "Kirsten Hein"
date: "2023-11-01"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### Load Packages and Set-up Environment
```{r}
library(sf)
library(sp)
library(rnaturalearthhires)
library(rnaturalearth)
library(RColorBrewer)
library(ggplot2)
library(viridis)
library(vegan)
library(raster)
library(terra)
library(pegas)

load("rda.RData") # load RDA results from vegan package

```
# *Current Adaptive Indices*

### Scale and Center Select Environmental Variables
```{r}
pred <- read.csv(env_subset_rda.csv")
pred <- pred[-1]
Env <- scale(pred, center = TRUE, scale = TRUE)
scale_env <- attr(Env, 'scaled:scale')
center_env <- attr(Env, 'scaled:center')
```

### Create "env_pres" raster file
```{r}
r <- raster::getData("worldclim", var="bio",res=10)
et <- ne_states(country = "ethiopia", returnclass = "sf")
geometry <- et$geometry
sp_geometry <- as(geometry, "Spatial")
raster_subset <- raster::mask(r, mask = sp_geometry)
```

### Creat function to project the adaptive component turnover across the landscape
```{r}
adaptive_index <- function(RDA, K, env_pres, range = NULL, method = "loadings", scale_env, center_env){
# Formatting environmental rasters for projection
var_env_proj_pres <- as.data.frame(rasterToPoints(env_pres[[row.names(RDA$CCA$biplot)]]))
# Standardization of the environmental variables
var_env_proj_RDA <- as.data.frame(scale(var_env_proj_pres[,-c(1,2)], center_env[row.names(RDA$CCA$biplot)], scale_env[row.names(RDA$CCA$biplot)]))
# Predicting pixels genetic component based on RDA axes
Proj_pres <- list()
if(method == "loadings"){
for(i in 1:K){
ras_pres <- rasterFromXYZ(data.frame(var_env_proj_pres[,c(1,2)], Z = as.vector(apply(var_env_proj_RDA[,names(RDA$CCA$biplot[,i])], 1, function(x) sum( x * RDA$CCA$biplot[,i])))), crs = crs(env_pres))
names(ras_pres) <- paste0("RDA_pres_", as.character(i))
Proj_pres[[i]] <- ras_pres
names(Proj_pres)[i] <- paste0("RDA", as.character(i))
}
}
# Prediction with RDA model and linear combinations
if(method == "predict"){
pred <- predict(RDA, var_env_proj_RDA[,names(RDA$CCA$biplot[,i])], type = "lc")
for(i in 1:K){
ras_pres <- rasterFromXYZ(data.frame(var_env_proj_pres[,c(1,2)], Z = as.vector(pred[,i])), crs = crs(env_pres))
names(ras_pres) <- paste0("RDA_pres_", as.character(i))
Proj_pres[[i]] <- ras_pres
names(Proj_pres)[i] <- paste0("RDA", as.character(i))
}
}
# Mask with the range if supplied
if(!is.null(range)){
Proj_pres <- lapply(Proj_pres, function(x) mask(x, range))
}
# Returning projections for current climates for each RDA axis
return(Proj_pres = Proj_pres)
}
```

### Calculate Adaptive Indices for Current Conditions
```{r}
res_RDA_proj_current <- adaptive_index(RDA = RDA.rda, K = 2, env_pres = raster_subset, method = "loadings", scale_env = scale_env, center_env = center_env)
RDA_proj <- list(res_RDA_proj_current$RDA1, res_RDA_proj_current$RDA2)
RDA_proj <- lapply(RDA_proj, function(x) rasterToPoints(x))
for(i in 1:length(RDA_proj)){
RDA_proj[[i]][,3] <- (RDA_proj[[i]][,3]-min(RDA_proj[[i]][,3]))/(max(RDA_proj[[i]][,3])-min(RDA_proj[[i]][,3]))
}
TAB_RDA <- as.data.frame(do.call(rbind, RDA_proj[1:2]))
colnames(TAB_RDA)[3] <- "value"
TAB_RDA$variable <- factor(c(rep("RDA1", nrow(RDA_proj[[1]])), rep("RDA2", nrow(RDA_proj[[2]]))), levels = c("RDA1","RDA2"))
```

### Plot Current Projection
```{r}
ggplot(data = TAB_RDA) +
geom_sf(data = et, fill=gray(.9), size=0) +
geom_raster(aes(x = x, y = y, fill = cut(value, breaks=seq(0, 1, length.out=10), include.lowest = T))) +
scale_fill_viridis_d(alpha = 0.8, direction = -1, option = "A", labels = c("Negative scores","","","","Intermediate scores","","","","Positive scores")) +
geom_sf(data = et, fill=NA, size=0.1) +
coord_sf(xlim = c(33, 48), ylim = c(3, 15), expand = FALSE) +
xlab("Longitude") + ylab("Latitude") +
guides(fill=guide_legend(title="Adaptive index")) +
facet_grid(~ variable) +
theme_bw(base_size = 11, base_family = "Times") +
theme(panel.grid = element_blank(), plot.background = element_blank(), panel.background = element_blank(), strip.text = element_text(size=11))
```
#######################################################################################################################################################################
### RDA can be used to predict the optimal adaptive index for each environmental pixel under consideration, using both current and future environmental conditions. The difference between the two predictions provides an estimate of the shift in adaptive index that would be required to track climate change. This proxy has been successively named genetic/genomic offset (Fitzpatrick & Keller, 2015), risk of non-adaptedness (Rellstab et al., 2016) or genomic vulnerability (Bay et al., 2018).

# Begin by downloading the ensemble projections for 2080.

# Go to this site: https://adaptwest.databasin.org/pages/adaptwest-climatena-cmip5/

# Scroll down to the following data sets and download them:
# *** In all cases, download the ASCII format ***
  # (1) Projection, Ensemble of 13 CMIP6 AOGCMs, RCP 4.5, SSP2, 2080

# Unzip using the default folder names & put these into a folder called "ClimateNA" located in the folder "Data"
#######################################################################################################################################################################

# *Projected Adaptive Indices*

### Load ClimateNA bioclimatic variables predicted for 2080 using RCP4.5 SSP2 climate scenario and the ensemble projection of 13 CMIP5 AOGCMs:
```{r}
# Get a list of GEOTIFF files in the specified directory
geotiff_files <- list.files("./", pattern = ".tif$", full.names = TRUE) #run outside of chunk

# Read GEOTIFF files into a raster stack
ras_2080 <- stack(geotiff_files)

# Extract variable names from file names
names(ras_2080) <- gsub(".tif$", "", tools::file_path_sans_ext(basename(geotiff_files)))

# Calculate ensemble mean for each bioclimatic variable
ensemble_mean <- calc(ras_2080, fun = mean, na.rm = TRUE)

# Save the ensemble mean as a GEOTIFF file
writeRaster(ensemble_mean, filename = "./ensemble_mean_2080.tif", format = "GTiff", overwrite = TRUE)
```

### Create RasterStack for each GCM
```{r}
# Get a list of GEOTIFF files in the specified directory
geotiff_files <- list.files("./", pattern = "wc*.tif$", full.names = TRUE)
names_list <- c("AC","CM","EC","FI","GI","HA","IN","IP","MI","MP","MR","UK") # this can change depending on the chosen models

# Create RasterStacks

for(i in 1:length(names_list) ){
  name <- lapply(names_list, function(names) names_list[[i]])
  raster <- stack(geotiff_files[i])
  names(raster) <-paste0("bio", 1:19)
  assign(names_list[i], raster)
}

# Extract 19 bioclimatic layers from each GCM RasterStack and Save

for(j in names_list){
    paste(print(j))
    raster_stack <- get(j)
    for(i in 1:nlayers(raster_stack)){
        band <- raster_stack[[i]]
        print(band)
        writeRaster(band,paste(print(j),'_',i,'.tif', sep=''), format = "Gtiff", overwrite = TRUE)
        
    }}

# Calculate Ensemble Means Across GCMs for each bioclimatic variable
### Change number in pattern and number in for-loop to match bio suffixes (e.g., 1:19). For consecutive runs, input one variable at a time. ###

geotiff_files <- list.files("./", pattern = "*_1.tif$", full.names = TRUE)
s <- stack(geotiff_files)

for(i in 1:19){
  geotiff_files <- list.files("./", pattern = paste("*_",i,".tif$",sep=''), full.names = TRUE)
  s <- stack(geotiff_files)
  emean <- calc(s, fun = mean, na.rm = TRUE)
  writeRaster(emean,paste('ensemble_mean_',i,'.tif', sep=''), format = "Gtiff", overwrite = TRUE)
}
```
#######################################################################################################################################################################

# *Genomic Offset*

### Function to predict genomic offset from an RDA model
```{r}
genomic_offset <- function(RDA, K, env_pres, env_fut, range = NULL, method = "loadings", scale_env, center_env){
  
  # Mask with the range if supplied
  if(!is.null(range)){
    env_pres <- mask(env_pres, range)
    env_fut <- mask(env_fut, range)
  }
  
  # Formatting and scaling environmental rasters for projection
  var_env_proj_pres <- as.data.frame(scale(rasterToPoints(env_pres[[row.names(RDA$CCA$biplot)]])[,-c(1,2)], center_env[row.names(RDA$CCA$biplot)], scale_env[row.names(RDA$CCA$biplot)]))
  var_env_proj_fut <- as.data.frame(scale(rasterToPoints(env_fut[[row.names(RDA$CCA$biplot)]])[,-c(1,2)], center_env[row.names(RDA$CCA$biplot)], scale_env[row.names(RDA$CCA$biplot)]))

  # Predicting pixels genetic component based on the loadings of the variables
  if(method == "loadings"){
    # Projection for each RDA axis
    Proj_pres <- list()
    Proj_fut <- list()
    Proj_offset <- list()
    for(i in 1:K){
      # Current climates
      ras_pres <- env_pres[[1]]
      ras_pres[!is.na(ras_pres)] <- as.vector(apply(var_env_proj_pres[,names(RDA$CCA$biplot[,i])], 1, function(x) sum( x * RDA$CCA$biplot[,i])))
      names(ras_pres) <- paste0("RDA_pres_", as.character(i))
      Proj_pres[[i]] <- ras_pres
      names(Proj_pres)[i] <- paste0("RDA", as.character(i))
      # Future climates
      ras_fut <- env_fut[[1]]
      ras_fut[!is.na(ras_fut)] <- as.vector(apply(var_env_proj_fut[,names(RDA$CCA$biplot[,i])], 1, function(x) sum( x * RDA$CCA$biplot[,i])))
      Proj_fut[[i]] <- ras_fut
      names(ras_fut) <- paste0("RDA_fut_", as.character(i))
      names(Proj_fut)[i] <- paste0("RDA", as.character(i))
      # Single axis genetic offset 
      Proj_offset[[i]] <- abs(Proj_pres[[i]] - Proj_fut[[i]])
      names(Proj_offset)[i] <- paste0("RDA", as.character(i))
    }
  }
  
  # Predicting pixels genetic component based on predict.RDA
  if(method == "predict"){ 
    # Prediction with the RDA model and both set of envionments 
    pred_pres <- predict(RDA, var_env_proj_pres[,-c(1,2)], type = "lc")
    pred_fut <- predict(RDA, var_env_proj_fut[,-c(1,2)], type = "lc")
    
    # List format
    Proj_offset <- list()    
    Proj_pres <- list()
    Proj_fut <- list()
    for(i in 1:K){
      # Current climates
      ras_pres <- rasterFromXYZ(data.frame(var_env_proj_pres[,c(1,2)], Z = as.vector(pred_pres[,i])), crs = crs(env_pres))
      names(ras_pres) <- paste0("RDA_pres_", as.character(i))
      Proj_pres[[i]] <- ras_pres
      names(Proj_pres)[i] <- paste0("RDA", as.character(i))
      # Future climates
      ras_fut <- rasterFromXYZ(data.frame(var_env_proj_pres[,c(1,2)], Z = as.vector(pred_fut[,i])), crs = crs(env_pres))
      names(ras_fut) <- paste0("RDA_fut_", as.character(i))
      Proj_fut[[i]] <- ras_fut
      names(Proj_fut)[i] <- paste0("RDA", as.character(i))
      # Single axis genetic offset 
      Proj_offset[[i]] <- abs(Proj_pres[[i]] - Proj_fut[[i]])
      names(Proj_offset)[i] <- paste0("RDA", as.character(i))
    }
  }
  
  # Weights based on axis eigen values
  weights <- RDA$CCA$eig/sum(RDA$CCA$eig)
  
  # Weighing the current and future adaptive indices based on the eigen values of the associated axes
  Proj_offset_pres <- do.call(cbind, lapply(1:K, function(x) rasterToPoints(Proj_pres[[x]])[,-c(1,2)]))
  Proj_offset_pres <- as.data.frame(do.call(cbind, lapply(1:K, function(x) Proj_offset_pres[,x]*weights[x])))
  Proj_offset_fut <- do.call(cbind, lapply(1:K, function(x) rasterToPoints(Proj_fut[[x]])[,-c(1,2)]))
  Proj_offset_fut <- as.data.frame(do.call(cbind, lapply(1:K, function(x) Proj_offset_fut[,x]*weights[x])))
  
  # Predict a global genetic offset, incorporating the K first axes weighted by their eigen values
  ras <- Proj_offset[[1]]
  ras[!is.na(ras)] <- unlist(lapply(1:nrow(Proj_offset_pres), function(x) dist(rbind(Proj_offset_pres[x,], Proj_offset_fut[x,]), method = "euclidean")))
  names(ras) <- "Global_offset"
  Proj_offset_global <- ras
  
  # Return projections for current and future climates for each RDA axis, prediction of genetic offset for each RDA axis and a global genetic offset 
  return(list(Proj_pres = Proj_pres, Proj_fut = Proj_fut, Proj_offset = Proj_offset, Proj_offset_global = Proj_offset_global, weights = weights[1:K]))
}
```

### Load Ensemble Files and Calculate Genomic Offset
```{r}

e_files <- list.files("./", pattern = "ensemble_*", full.names = T)

ras_2080 <- stack(e_files)

names(ras_2080) <- c("bio1","bio10","bio11","bio12","bio13","bio14","bio15","bio16","bio17","bio18","bio19","bio2","bio3","bio4","bio5","bio6","bio7","bio8","bio9")

ras_2080_crop <- raster::mask(ras_2080, mask = sp_geometry)


res_RDA_proj2080 <- genomic_offset(wolf.rda, K = 2, env_pres = raster_subset, env_fut = ras_2080_crop, range = NULL, method = "loadings", scale_env = scale_env, center_env = center_env)
``` 

### Calculate Global Genomic Offset
```{r}
RDA_proj_offset <- data.frame(rasterToPoints(res_RDA_proj2080$Proj_offset_global))
RDA_proj_offset$Date <- rep("2080", nrow(rasterToPoints(res_RDA_proj2080$Proj_offset_global)))
```

### Plot Global Genomic Offset
```{r}
colors <- c(colorRampPalette(brewer.pal(11, "Spectral")[6:5])(3), colorRampPalette(brewer.pal(11, "Spectral")[4:3])(4), colorRampPalette(brewer.pal(11, "Spectral")[2:1])(5))

ggplot(data = RDA_proj_offset) + 
    geom_sf(data = et, fill=gray(.9), size=0) +
    geom_raster(aes(x = x, y = y, fill = cut(Global_offset, breaks=seq(0, 2, by = 0.09), include.lowest = T)), alpha = 1) + 
    scale_fill_manual(values = colors, labels = c("0.5","","","","1.0","","","","", "1.5"), guide = guide_legend(title="Genomic offset", title.position = "top", title.hjust = 0.5, ncol = 1, label.position="right"), na.translate = F) +
    geom_sf(data = et, fill=NA, size=0.1) +
    coord_sf(xlim = c(33, 48), ylim = c(3, 15), expand = FALSE) +
    xlab("Longitude") + ylab("Latitude") +
    facet_grid(~ Date) +
    theme_bw(base_size = 11, base_family = "Times") +
    theme(panel.grid = element_blank(), plot.background = element_blank(), panel.background = element_blank(), strip.text = element_text(size=11))
```

## Plotting Present - Future Climate Projections + Genomic Offsets

```{r}
### Prepare Current Projections Table
RDA_proj <- list(res_RDA_proj2080$Proj_pres$RDA1, res_RDA_proj2080$Proj_pres$RDA2)
RDA_proj <- lapply(RDA_proj, function(x) rasterToPoints(x))
for(i in 1:length(RDA_proj)){
     RDA_proj[[i]][,3] <- (RDA_proj[[i]][,3]-min(RDA_proj[[i]][,3]))/(max(RDA_proj[[i]][,3])-min(RDA_proj[[i]][,3]))
 }
 TAB_RDA <- as.data.frame(do.call(rbind, RDA_proj[1:2]))
 colnames(TAB_RDA)[3] <- "value"
 TAB_RDA$variable <- factor(c(rep("RDA1", nrow(RDA_proj[[1]])), rep("RDA2", nrow(RDA_proj[[2]]))), levels = c("RDA1","RDA2"))
 
### Prepare Future Projections Table
RDA_proj2 <- list(res_RDA_proj2080$Proj_fut$RDA1, res_RDA_proj2080$Proj_fut$RDA2)
RDA_proj2 <- lapply(RDA_proj2, function(x) rasterToPoints(x))
for(i in 1:length(RDA_proj2)){
     RDA_proj2[[i]][,3] <- (RDA_proj2[[i]][,3]-min(RDA_proj2[[i]][,3]))/(max(RDA_proj2[[i]][,3])-min(RDA_proj2[[i]][,3]))
 }
 PRED_RDA <- as.data.frame(do.call(rbind, RDA_proj2[1:2]))
 colnames(PRED_RDA)[3] <- "value"
 PRED_RDA$variable <- factor(c(rep("RDA1", nrow(RDA_proj2[[1]])), rep("RDA2", nrow(RDA_proj2[[2]]))), levels = c("RDA1","RDA2"))
```

### Prepare Genomic Offsets Table  
```{r}
RDA_proj3 <- list(res_RDA_proj2080$Proj_offset$RDA1, res_RDA_proj2080$Proj_offset$RDA2)
RDA_proj3 <- lapply(RDA_proj3, function(x) rasterToPoints(x))
for(i in 1:length(RDA_proj3)){
     RDA_proj3[[i]][,3] <- (RDA_proj3[[i]][,3]-min(RDA_proj3[[i]][,3]))/(max(RDA_proj3[[i]][,3])-min(RDA_proj3[[i]][,3]))
 }
 OFF_RDA <- as.data.frame(do.call(rbind, RDA_proj2[1:2]))
 colnames(OFF_RDA)[3] <- "value"
 OFF_RDA$variable <- factor(c(rep("RDA1", nrow(RDA_proj3[[1]])), rep("RDA2", nrow(RDA_proj3[[2]]))), levels = c("RDA1","RDA2"))

### Prepare Current and Future Projections Tables for GGPLOT2  
TAB_RDA$group <- "2018"
TAB_RDA$group <- as.factor(TAB_RDA$group)
PRED_RDA$group <- "2080"
PRED_RDA$group <- as.factor(PRED_RDA$group)
combined_RDA <- rbind(TAB_RDA,PRED_RDA)
```

### Plot Current and Future Projections onto Map
```{r}
ggplot(data = combined_RDA) +
    geom_sf(data = et, fill=gray(.9), size=0) +
    geom_raster(aes(x = x, y = y, fill = cut(value, breaks=seq(0, 1, length.out=10), include.lowest = T))) +
    scale_fill_viridis_d(alpha = 0.8, direction = -1, option = "A", labels = c("Negative scores","","","","Intermediate scores","","","","Positive scores")) +
    geom_sf(data = et, fill=NA, size=0.1) +
    coord_sf(xlim = c(33, 48), ylim = c(3, 15), expand = FALSE) +
    xlab("Longitude") + ylab("Latitude") +
    guides(fill=guide_legend(title="Adaptive index")) +
    facet_grid(variable ~ group) +
    theme_bw(base_size = 11, base_family = "Times") +
    theme(panel.grid = element_blank(), plot.background = element_blank(), panel.background = element_blank(), strip.text = element_text(size=11))
```

### Prepare Genomic Offsets Table for GGPLOT2
```{r}
OFF_RDA$group <- "Genomic Offset"
OFF_RDA$group <- as.factor(OFF_RDA$group)
colors <- c(colorRampPalette(brewer.pal(11, "Spectral")[6:5])(2), colorRampPalette(brewer.pal(11, "Spectral")[4:3])(2), colorRampPalette(brewer.pal(11, "Spectral")[2:1])(3))
```

### Plot Genomic Offsets Projections onto Map
```{r}
ggplot(data = OFF_RDA) +
    geom_sf(data = et, fill=gray(.9), size=0) +
    geom_raster(aes(x = x, y = y, fill = cut(value, breaks=seq(0, 1, length.out=5), include.lowest = T))) +
    scale_fill_manual(values = colors, guide = guide_legend(title="Genomic offset", title.position = "top", title.hjust = 0.5, ncol = 1, label.position="right"), na.translate = FALSE) +
    geom_sf(data = et, fill=NA, size=0.1) +
    coord_sf(xlim = c(33, 48), ylim = c(3, 15), expand = FALSE) +
    xlab("Longitude") + ylab("Latitude") +
    guides(fill=guide_legend(title="Genomic Offset")) +
    facet_grid(variable ~ group) +
    theme_bw(base_size = 11, base_family = "Times") +
    theme(panel.grid = element_blank(), plot.background = element_blank(), panel.background = element_blank(), strip.text = element_text(size=11))

ggplot(data = OFF_RDA) +
    geom_sf(data = et, fill=gray(.9), size=0) +
    geom_raster(aes(x = x, y = y, fill = cut(value, breaks=seq(0, 1, length.out=6), include.lowest = T))) +
    scale_fill_manual(values = colors, labels = c("0.0 - 0.2", "0.2 - 0.4","0.4 - 0.6","0.6 - 0.8","0.8 - 1.0"), guide = guide_legend(title="Genomic offset", title.position = "top", title.hjust = 0.5, ncol = 1, label.position="right"), na.translate = FALSE) +
    geom_sf(data = et, fill=NA, size=0.1) +
    coord_sf(xlim = c(33, 48), ylim = c(3, 15), expand = FALSE) +
    xlab("Longitude") + ylab("Latitude") +
    guides(fill=guide_legend(title="Genomic Offset")) +
    facet_grid(variable ~ group) +
    theme_bw(base_size = 11, base_family = "Times") +
    theme(panel.grid = element_blank(), plot.background = element_blank(), panel.background = element_blank(), strip.text = element_text(size=11))

```
