### Code was adapted from https://github.com/Yunqing-Yu/tef-abscission-zone/blob/main/ClusterProfiler-dabbi_final.R
### For Bioconductor Genome wide annotation of tef and use of code below, please contact and cite:
### Yunqing Yu; https://doi.org/10.1093/plphys/kiad079

---
title: "Gene Ontology Enrichment"
author: "Kirsten Hein"
date: "2023-12-05"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### Install SvOrgDb from local
```{r}
install.packages("/local/org.Etef.eg.db",repos = NULL, type="source")
```

### Load Packages and Set-up Environment
```{r}
library(clusterProfiler)
library(org.Etef.eg.db)
library(GOSemSim)
library(enrichplot)
library(writexl)
library(readxl)
library(dplyr)

```

### Load gene list from GO_input - Kirsten Hein Code
```{r}
gene_names <- read.table("/local/Gene_ID_clusterProfiler/GO_genelist.csv", quote="\"", comment.char="") # list of RDA candidate genes
gene_list <- gene_names$V1
gene_list <- as.list(gene_list)
```

### Make GOsemsimDATA object used for enrichment analysis 
```{r}
SemData_tef <- godata(OrgDb = "org.Etef.eg.db", keytype = "GID", ont = "BP", computeIC = TRUE)
```

######### Test parameters for enrichGO #########
```{r}
ego_s1 <- enrichGO(gene = gene_list[[1]], 
                   OrgDb = "org.Etef.eg.db", keyType="GID", ont="BP", 
                   pAdjustMethod = "BH", qvalueCutoff = 0.05, 
                   minGSSize = 10, maxGSSize = 1000)

# These methods are all similar but not exactly the same. I ended up deciding to use "Rel".
ego_s1_Rel <- simplify(ego_s1, cutoff = 0.6, measure = "Rel", semData = SemData_tef)
ego_s1_Rel@result$Description

ego_s1_Jiang <- simplify(ego_s1, cutoff = 0.6, measure = "Jiang", semData = SemData_tef)
ego_s1_Jiang@result$Description

ego_s1_Wang <- simplify(ego_s1, cutoff = 0.6, measure = "Wang", semData = SemData_tef)
ego_s1_Wang@result$Description

ego_s1_Lin <- simplify(ego_s1, cutoff = 0.6, measure = "Lin", semData = SemData_tef)
ego_s1_Lin@result$Description

ego_s1_Resnik <- simplify(ego_s1, cutoff = 0.6, measure = "Resnik", semData = SemData_tef)
ego_s1_Resnik@result$Description
```
######### Perform enrichGO analysis #########
### Use enricher for visualizing a single gene list
```{r}
ego_list <- list()  #store enrichGO output
ego_list_Rel <- list()   #store enrichGO output after simplify with "Rel" method
GO_list <- list()    #store result table
GO_list_Rel <- list()   #store result table after simplify with "Rel" method
updated_GO_list <- list()
updated_ego_list_Rel <- list()
updated_GO_list_Rel <- list()

for (module in seq_along(gene_list)){
  GENE_ID <- gene_list[[module]]
  ego <- enrichGO(gene = GENE_ID, 
                  OrgDb = "org.Etef.eg.db", keyType="GID", ont="BP", 
                  pAdjustMethod = "BH", qvalueCutoff = 0.05, 
                  minGSSize = 10, maxGSSize = 1000) 
  ego_list[[module]] <- ego
}

ego_list2 <- Filter(function(x) !is.null(x), ego_list)

for (module in seq_along(ego_list2)){
GO_list[[module]] <- ego_list2[[module]][ego_list2[[1]]$p.adjust < 0.05,]
gene_value <- ego_list2[[module]]@gene
updated_GO_list[[gene_value]] <- GO_list[[module]]
}

for (module in seq_along(ego_list2)){
ego_Rel <- simplify(ego_list2[[module]], cutoff = 0.6, measure = "Rel", semData = SemData_tef
                   
ego_list_Rel[[module]] <- ego_Rel
GO_list_Rel[[module]] <- ego_Rel@result
gene_value <- ego_list2[[module]]@gene
updated_ego_list_Rel[[gene_value]] <- ego_list_Rel[[module]]
updated_GO_list_Rel[[gene_value]] <- GO_list_Rel[[module]]
}

# Save files
write_xlsx(updated_GO_list, "GO_clusterProfiler.xlsx")
write_xlsx(updated_GO_list_Rel, "GO_clusterProfiler_Rel0.6.xlsx")
```
                    
### Create excel file of GO terms (unlisted) with environmental data
```{r}
# Load gene model (highest correlation) RDA environmental predictors
env <- read.csv("/local/Gene_ID_clusterProfiler/GO_geneID_RDApredictors.csv")

# Unlist GO terms      
combined_df <- data.frame()
for (i in seq_along(updated_GO_list_Rel)) {
  combined_df <- rbind(combined_df, updated_GO_list_Rel[[i]])}

# Retain distinct terms and merge with environmental data
combined_df2 <- combined_df %>% distinct()
combined_df2 <- merge(combined_df2, env, by = "geneID")

# Write excel file
write_xlsx(combined_df2, "GO_clusterProfiler_Rel0.6_unlisted.xlsx")
```
############################################################################################################################
### Visualization of GO enrichment analysis for 63 of the RDA-1kb genes. 
#### Analysis was limited to genes that had annotations in the SvOrgdb and had gene sets between 10 - 1000
#### Other gene set parameters were tested (lower and higher), no differences were observed.

### Dot Plots
```{r}
dot_dabbi <- lapply(names(updated_ego_list_Rel), function(x){dotplot(updated_ego_list_Rel[[x]], x = "count", showCategory=100, title = x, font.size =6, label_format = 30)})
dot_dabbi[[1]] # replace number with list number associated with GO term of interest
```

### Use compareCluster for comparison between gene lists
```{r}
# Cluster and compare all gene models (no environment)
Cluster_AZ <- compareCluster(names(updated_ego_list_Rel), fun = "enrichGO",
                             pAdjustMethod = "BH", qvalueCutoff = 0.05, 
                             minGSSize = 10, maxGSSize = 1000)

# Cluster and compare all gene models by environment
Cluster_AZ_env <- compareCluster(geneID ~ RDA_predictor, data = env, fun = "enrichGO",
                             OrgDb = "org.Etef.eg.db", keyType="GID", ont="BP", 
                             pAdjustMethod = "BH", qvalueCutoff = 0.05, 
                             minGSSize = 10, maxGSSize = 1000)

#Simplify redundant terms
Cluster_AZ_s <- simplify(Cluster_AZ, cutoff=0.6, by="p.adjust", measure = "Rel", semData = SemData_tef)
Cluster_AZ_s_env <- simplify(Cluster_AZ_env, cutoff=0.6, by="p.adjust", measure = "Rel", semData = SemData_tef)
```
                    
# compareCluster result visualization
```{r}
# Dot plots
dotplot(Cluster_AZ_s, showCategory = 10, font.size =8, label_format = 30)
dotplot(Cluster_AZ_s_env, showCategory = 10, font.size =8, label_format = 30)

# Mapping networks - visualization
pairwise_cluster_s_env <- pairwise_termsim(Cluster_AZ_s_env)
p1 <- emapplot(pairwise_cluster_s_env)
p1 + scale_fill_manual(values=c("darkorchid4","red","chartreuse4","yellow","slategray1" ))

p1 <- emapplot_cluster(pairwise_cluster_s_env, showCategory = 30)
p1 + scale_fill_manual(values=c("darkorchid4","red","chartreuse4","yellow","slategray1" ))
treeplot(pairwise_cluster_s_env, fontsize = 3, cex_category = .1, offset.params = list(extend = .1), mar = c(5, 2, 5, 5), align = "right", hexpand = 0.5, offset = rel(.75), offset_tiplab = rel(1.5)) + theme(axis.text.x = element_text(size = 3)) 

# Tree diagrams
treeplot(pairwise_cluster_s_env, fontsize = 3, cex_category = .1, offset.params = list(extend = .1), mar = c(5, 2, 5, 5), align = "right", hexpand = 0.5, offset = rel(.75), offset_tiplab = rel(1.5)) + theme(axis.text.x = element_text(size = 3))
```

```                    
                    
